---
title: "How to use PropensitySub"  
output: rmarkdown::html_vignette 
vignette: >
  %\VignetteIndexEntry{How to use PropensitySub}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

In biomedical research, biomarker is an important tool to characterize subpopulations with better/worse disease progression, or different response to a treatment/intervention.
When comparing two different treatments, to estimate the treatment benefit within a biomarker subpopulation, ideally the biomarker should be collected in two comparative cohorts with the same procedure - for example, collecting biomarker as a baseline stratification factor in a randomized trial.

However, there are a number of cases where it is impossible to collect biomarker samples from both cohorts. For example, the biomarker positive/negative status is only available in the treatment cohort but not observable in the control cohort. In this case, if the treatment benefit in the biomarker positive group is of interest, one needs to compare the biomarker positive group from the treatment cohort against the “unobserved” biomarker positive group from the control cohort. The “unobserved” biomarker positive group is the subjects who are likely to be biomarker positive if the biomarker is measurable. To predict the “unobserved” biomarker positive group, one may predict the hidden status based on other observed associated features.

This package implements multiple approaches for this type of modeling. In addition to the modeling (session 1), the package also provides handy functions on model fitting diagnostics (session 3 and 4), visualization (session 2, 5), and summarization (session 6).


# 0. Example Data: dataset "biomarker" from package `PropensitySub`
 
  - Endpoint: Overall survival (`OS`, `OS.event`)
  - Treatment Group: "Experimental" and "Control" (`Arm`)
  - Strata: "Positive" and "Negative" (`STRATUM`)
  
 
```{r, message=FALSE, warning=FALSE, comment=""}
library(PropensitySub) 
library(dplyr)
head(biomarker, 3)
#  Control arm subjects have no STRATUM measurments
#  Experimental arm subjects partially miss STRATUM measurement.
table(biomarker$Arm, biomarker$STRATUM, useNA = "ifany")
```

# 1. Estimate treatment effect within each `STRATUM`.



## 1.1 Data process: generate a numeric version of strata labeling


Below shows example of 
1) imputing missing as negative (assuming missing not at random, and evaluating an extreme case where all missing data are supposed to be negative); 2) treating missing as a separate class: also assuming missing not at random.

`indicator_1`: numeric version of `STRATUM` where missing in Experimental Arm is imputed as "Negative".   
`indicator_2`: numeric version of `STRATUM` where missing is kept as a seperate stratum 
```{r, message=FALSE}
biomarker <- biomarker %>%
  mutate(
    indicator_1 = case_when(
      STRATUM == "Positive" ~ 1,
      STRATUM == "Negative" ~ 0,
      # impute missing as "Negative" in Experimental Arm
      Arm == "Experimental" & is.na(STRATUM) ~ 0,
      is.na(STRATUM) & Arm == "Control" ~ -1
    ),
    indicator_2 = case_when(
      STRATUM == "Positive" ~ 1,
      STRATUM == "Negative" ~ 0,
      # keep missing as a seperate stratum in Experimental Arm
      Arm == "Experimental" & is.na(STRATUM) ~ 2,
      is.na(STRATUM) & Arm == "Control" ~ -1
    ),
    # treatment group needs to be factor
    Arm = factor(Arm, levels = c("Control", "Experimental"))
  
  )
```

## 1.2 Models for analyzing treatment effect


For subjects in the control arm, to predict a subject’s likelihood of being in different biomarker strata, the first step is calculating P(A | X), where A indicates the biomarker strata and X indicates observed associated features. The model may be learned using the treatment data where both biomarker strata and associated observed features were measured, and be applied in control cohort data to predict the biomarker strata assignments. 

Once P(A|X) is calculated, predicted biomarker group of interest in the control cohort can be obtained by inverse probability weighting (IPW) or propensity score matching (PSM). Both IPW and PSM are implemented in the package.


In IPW, the predicted probability of being stratum A will be used as weights when estimating treatment difference of two arms (Hazard ratio for survival endpoint; response rate difference for binary endpoint).

In PSM, the predicted probability of being stratum A will be used for propensity score matching.
The matching results will then be used to estimate treatment difference of two arms (Hazard ratio for survival endpoint; response rate difference for binary endpoint).

Below shows different modeling approaches implemented in this package.

### 1.2.1 `plain`: binomial/multinomial `glm`  
```{r, message=FALSE, comment=""}
# `plain` model with IPW method and aggressive imputation where missing is imputated as negative
ipw_plain_str2 <- ipw_strata(
  data.in = biomarker, formula = indicator_1 ~ ECOG + Sex + Age, model = "plain",
  indicator.var = "indicator_1", tte = "OS", event = "OS.event", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0)
)
# get weighted HRs
ipw_plain_str2$stat
# check model converge
ipw_plain_str2$converged
# get weights
ipw_plain_str2$data %>% 
  dplyr::select(Patient.ID, Arm, STRATUM, ECOG, Sex, Age, indicator_1, pred1, pred0) %>%
  head()


# `plain` model with IPW method and missing is kept as another stratum
ipw_plain_str3 <- ipw_strata(
  data.in = biomarker, formula = indicator_2 ~ ECOG + Sex + Age, model = "plain",
  indicator.var = "indicator_2", tte = "OS", event = "OS.event", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0, "Unknown" = 2)
)
# get weighted HRs
ipw_plain_str3$stat

# `plain` model with PSM method and aggressive imputation
ps_plain_str2 <- ps_match_strata(
  data.in = biomarker, formula = indicator_1 ~ ECOG + Sex + Age, model = "plain",
  indicator.var = "indicator_1", tte = "OS", event = "OS.event", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0)
)
# get weighted HRs
ps_plain_str2$stat 
```
### 1.2.2 `dwc`: doubly weighted control
```{r, message=FALSE, comment=""}
# dwc model with IPW method
ipw_dwc <- ipw_strata(
  data.in = biomarker, formula = indicator_2 ~ ECOG + Sex + Age, model = "dwc",
  indicator.var = "indicator_2", tte = "OS", event = "OS.event", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0)
)
# get weighted HRs
ipw_dwc$stat 

# dwc model with PSM method
ps_dwc <- ps_match_strata(
  data.in = biomarker, formula = indicator_2 ~ ECOG + Sex + Age, model = "dwc",
  indicator.var = "indicator_2", tte = "OS", event = "OS.event", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0, "Missing" = 2)
)
# get weighted HRs
ps_dwc$stat 
```

### 1.2.3 `wri`: weight regression imputation, currently only available with IPW method

```{r, message=FALSE, comment=""}
# data process: create a numeric version of next STRATUM to learn from  
biomarker <- biomarker %>%
  mutate(
    indicator_next_2 = case_when(
      STRATUM.next == "Positive" ~ 1,
      STRATUM.next == "Negative" ~ 0, 
      Arm == "Experimental" & is.na(STRATUM.next) ~ 2,
      is.na(STRATUM.next) & Arm == "Control" ~ -1
    )
  )
ipw_wri <- ipw_strata(
  data.in = biomarker, formula = indicator_2 ~ ECOG + Sex + Age, model = "wri",
  indicator.var = "indicator_2", indicator.next = "indicator_next_2",
  tte = "OS", event = "OS.event", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0)
)
# get weighted HRs
ipw_wri$stat 
```

 
# 2. Weighted KM plots from modeling

The treatment benefit within biomarker positive stratum can be calculated as the hazard ratio between the observed biomarker positive group in the treatment cohort against the propensity score matched biomarker positive group in the control cohort. An associated KM can be generated as well.  When IPW is used, P(A|X) are used as weights for the HR calculation and KM curves. Binary endpoint is also supported.


```{r, message=FALSE, fig.width=6, fig.height=5, comment=""}
# for ipw_plain model results
km_plot_weight(
  data.in = ipw_plain_str2$data,
  indicator.var = "indicator_1", tte = "OS", event = "OS.event", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0)
)

# to get weights from model result for further usage
ipw_plain_str2$data %>% 
  dplyr::select(Patient.ID, Arm, STRATUM, ECOG, Sex, Age, indicator_1, pred1, pred0) %>%
  head()
# for ipw_wri model results
km_plot_weight(
  data.in = ipw_wri$data,
  indicator.var = "indicator_2", tte = "OS", event = "OS.event", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0)
)
# for ps_dwc model results
km_plot_weight(
  data.in = ps_dwc$data,
  indicator.var = "indicator_2", tte = "OS", event = "OS.event", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0, "Missing" = 2)
)

```

# 3. Check standardized difference of covariates

The package also implemented the absolute standardized mean difference (ASMD) calculation to assess balance of the observed associated features (Xs)(Austin and Stuart, 2015). With a good model fitting, after IPW or PSM adjustment, one would expect a small number of features’ or no feature’s ASMD exceed certain thresholds (such as 0.1 and 0.25). One would expect to see smaller values comparing after-adjustment ASMDs to pre-adjustment ASMDs.

Theoretical calculation is also implemented. The theoretical calculation provides the expected number of features with ASMD greater than a threshold. The calculation depends on sample size and the number of features in X. One may compare the empirical number of features whose ASMD beyond a threshold to the theoretical calculation to assess model performance.

## 3.1 Example for `ipw` method with `plain` model
```{r, message=FALSE, fig.width=6, fig.height=5, comment=""}
ipw_plain_diff <- std_diff(
  data.in = ipw_plain_str2$data, vars = c("ECOG", "Sex", "Age"),
  indicator.var = "indicator_1", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0),
  usubjid.var = "Patient.ID"
)
ipw_plain_diff$Positive
ipw_plain_diff$Negative
# Visualize differences 
std_diff_plot(ipw_plain_diff, legend.pos = "bottom")

```

## 3.2 Example for `psm` method with `dwc` model

```{r, message=FALSE, fig.width=6, fig.height=5, comment=""}

ps_dwc_diff <- std_diff(
  data.in = ps_dwc$data, vars = c("ECOG", "Sex", "Age"),
  indicator.var = "indicator_2", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0, "Missing" = 2),
  usubjid.var = "Patient.ID"
)
ps_dwc_diff$Missing
 
# Visualize differences 
std_diff_plot(ps_dwc_diff)
```

# 4. Expected vs. Observed tables for imbalanced covariates to check model fit
```{r, message=FALSE, comment=""}
vars <- c("ECOG", "Sex", "Age")
thresholds <- c(0.15, 0.2)
class_int_list <- list("Positive" = 1, "Negative" = 0)
rand_ratio <- 2 # randomization ratio
n_arms <- lapply(class_int_list, function(x) nrow(biomarker %>% filter(indicator_1 %in% x)))
exp_diff <- sapply(n_arms, function(x) {
  expected_feature_diff(
    n.feature = length(vars),
    n.arm1 = x,
    n.arm2 = x / rand_ratio,
    threshold = thresholds
  )
}) %>% t()

# Expected imbalanced features
exp_diff

# Calculate the observed imbalanced features in model ipw_plain_diff
obs_diff_cnt <- sapply(thresholds, function(th){
  sapply(ipw_plain_diff, function(gp){
    ft <- as.character(subset(gp, type=="adjusted difference" & absolute_std_diff>=th)$var)
    length(unique(sapply(ft, function(ff)strsplit(ff, split="\\.")[[1]][1])))
  })
})
colnames(obs_diff_cnt) <- paste0("Observed # features > ", thresholds)
rownames(obs_diff_cnt) <- names(ipw_plain_diff)
# the number of observed imbalanced features for each threshold
# Compare expected to observed # of imbalanced features to check model fit
obs_diff_cnt

obs_diff_fac <- sapply(thresholds, function(th) {
  sapply(ipw_plain_diff, function(gp) {
    ft <- as.character(subset(gp, type=="adjusted difference" & absolute_std_diff>=th)$var)
    paste(ft, collapse=",")
  })
})
colnames(obs_diff_fac) <- paste0("features > ", thresholds)
rownames(obs_diff_fac) <- names(ipw_plain_diff)
# the observed individual features that are imbalanced for each threshold
obs_diff_fac


```

# 5. Bootstrap to get a robust CIs for the treatment effect

## 5.1 Example for `ipw` method with `plain` model

```{r, message=FALSE, comment=""}
boot_ipw_plain <- bootstrap_propen(
  data.in = biomarker, formula = indicator_1 ~ ECOG + Sex + Age,
  indicator.var = "indicator_1", tte = "OS", event = "OS.event", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0),
  estimate.res = ipw_plain_str2, method = "ipw", n.boot = 100
)
# get bootstrap CI
boot_ipw_plain$est.ci.mat
# summary statistics from bootstraps
boot_ipw_plain$boot.out.est
# error status and convergence status
boot_ipw_plain$error.est
boot_ipw_plain$conv.est
```

## 5.2 Example for `ipw` method with `wri` model

```{r, message=FALSE, comment=""}
boot_ipw_wri <- bootstrap_propen(
  data.in = biomarker, formula = indicator_2 ~ ECOG + Sex + Age,
  indicator.var = "indicator_2", indicator.next = "indicator_next_2",
  tte = "OS", event = "OS.event", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0),
  estimate.res = ipw_wri, method = "ipw", n.boot = 100
)
# get bootstrap CI
boot_ipw_wri$est.ci.mat  
```

## 5.3 Example for `psm` method with `dwc` model
```{r, message=FALSE, comment=""}
boot_ps_dwc <- bootstrap_propen(
  data.in = biomarker, formula = indicator_2 ~ ECOG + Sex + Age,
  indicator.var = "indicator_2",  
  tte = "OS", event = "OS.event", trt = "Arm",
  class.of.int = list("Positive" = 1, "Negative" = 0),
  estimate.res = ps_dwc, method = "ps", n.boot = 100
)
# get bootstrap CI
boot_ps_dwc$est.ci.mat  
```

# 6. Create forestplot to visualize multiple model results 
```{r, message=FALSE, fig.width=7, fig.height=6, comment=""}
boot_models <- list(
  ipw_plain = boot_ipw_plain, 
  ipw_wri = boot_ipw_wri, 
  ps_dwc = boot_ps_dwc
)
cols <- c("Estimate", "Bootstrap CI low", "Bootstrap CI high")
# get HRs and bootstrap CIs
boots_dp <- lapply(boot_models, function(x){
  cis <- x$est.ci.mat[ , cols, drop = FALSE] %>%
    exp() %>% round(2) %>% as.data.frame() 
  colnames(cis) <- c("HR", "LOWER", "UPPER")
  cis %>% mutate(
    Group = rownames(cis)
  )
})
boots_dp <- do.call(`rbind`, boots_dp) %>%
  mutate(
    Methods = rep(c("IPW plain", "IPW wri", "PS dwc"), 2),
    Methods_Group = paste(Methods, Group), 
    Group = factor(Group, levels = c("Positive", "Negative")),
    Methods = factor(Methods, levels = c("IPW plain", "IPW wri", "PS dwc"))
  ) %>%
  arrange(Methods, Group)
forest_bygroup(
  data = boots_dp, summarystat = "HR", upperci = "UPPER", lowerci = "LOWER",
  population.name = "Methods_Group", group.name = "Methods",
  color.group.name = "Group", 
  stat.label = "Hazard Ratio", 
  stat.type.hr = TRUE, log.scale = FALSE,  
  endpoint.name = "OS", study.name = "Example Study", draw = TRUE
)
```

# 7. Functions

Below are the main functions in this package :

- `ipw_strata()`: IPW, allows for 2 or multiple classes. It also allows for predicting both patients with unknown class label in control arm and patients with unknown class label in trt arm. Supports both survival endpoint (HR as effect size estimator) and response endpoint (ORR difference as effect size estimator)

- `ps_match_strata()`: PSM, similar to `ipw_strata()`

- `km_plot_weight()`: weighted km, can take outputs from either `ipw_strata()` or `ps_match_strata()`

- `bootstrap_propen()`: bootstrap function; can take outputs from both `ipw_strata()` and `ps_match_strata()`; outputs bootstrap CI as well as summary statistics (median, mean, etc) of bootstrap HRs (ORR deltas)

- `forest_bygroup()`: forestplot function; additional functionality to impose extra summary statistics on forestplot (e.g. median bootstrap HR on top of HR point estimate) 

- `std_diff()`: Calculate feature's standardized difference across two arms. This function can be used to check covariate balance between the (matched) strata from two arms.

- `expected_feature_diff()`: Calculate expected number of features showing standardized difference > threshold. This function can be used to assess whether the feature balance in the matched/adjusted data set is comparable to an RCT.
